Recursive partitioning of linear growth curve models

### 2022-11-11 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
move second paper to lgcm-tree folder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3236 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 872e6824c6a542582e905032ec1338bfa6c58b9c

\abstract{Growth curve models are popular tools for studying {\color{red!80} the development of a response variable within the same observational units} over time. Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time. Recursive partitioning methods (RPMs) allow for detecting such subgroups. In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning linear growth curve models (LGCMs). GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data. {\color{red!80} Here, we introduce an extension of the algorithm for longitudinal (as opposed to clustered) data that can detect subgroups defined by subject-specific covariates on/or time-varying covariates. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.} Using simulated and empirical data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart. We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for the analysis of trajectories with both discrete and continuous time, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

>>>>

\abstract{Growth curve models are popular tools for studying {\color{red!80} the development of a response variable within the same observational units} over time. Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time. Recursive partitioning methods (RPMs) allow for detecting such subgroups. In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning linear growth curve models (LGCMs). GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data. {\color{red!80} Here, we introduce an extension of the algorithm for longitudinal (as opposed to clustered) data that can detect subgroups defined by subject-specific covariates and/or time-varying covariates. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.} Using simulated and empirical data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart. We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for the analysis of trajectories with both discrete and continuous time, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}
